---
title: "Chapter 13: Sparse linear models"
author: "Jiali Lin"
date: "5/3/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(SML)
library(mvtnorm)
library(pscl)
library(latex2exp)
library(ggplot2)
library(reshape2)
library(R.matlab)
library(glmnet)
```

## 1. Example: Bayesian variables selection with continuous gaussian responses by gibbs sampling.

Let's generate and plot some synthetic data.

```{r}
set.seed(50)
sigma2 <- 4
p <- 5 
n <- 100 
true.gamma <- c(1, 0, 1, 0, 1)  
true.beta <- c(10, 0, 8, 0, 6)   
covariate <- matrix(rnorm(n * p, 0, 1), ncol = p, byrow = F) 
Y <- covariate %*% true.beta + rnorm(n, 0, sqrt(sigma2)) 
```

Rough estimate without initial values.

```{r}
X0 <- covariate
beta.hat <- solve(t(X0) %*% X0) %*% t(X0) %*% Y
sig2.hat <- sum((Y - X0 %*% beta.hat) ** 2) / n
```

Set inital values and parameters for MCMC.

```{r}
nMc <- 100
a <- (sig2.hat) ** 2 / 10 + 2 
b <- sig2.hat * (a - 1)
v0 <- rep(10e-3, p)
v1 <- rep(sqrt(max(beta.hat) * 100), p)
w <- 1 - sum(abs(beta.hat) < 4) / p
gamma0 <- rbinom(p, size = 1, prob = w)
sigma0 <- sig2.hat
```

Run the algorithm and estimate compuation time.

```{r}
## Tic
ptm <- proc.time()

## Run MCMC
res <- BvsGibbs(covariate, Y, nMc, v1, v0, w, gamma0, sigma0, a, b)

## Tok
Time.used <- proc.time() - ptm
Time.used[3] / 60
```

Get posterior mean of $\boldsymbol\beta$ and make plot.

```{r}
burnin <- 30
beta.sample <- res$sample.beta[(burnin + 1):nMc, ]
beta.mean <- colMeans(beta.sample)
par(mfrow <- c(1, 1))

qplot(x = 1:5, y = true.beta, colour = "Truth", geom = "point") + 
geom_point(aes(y = beta.mean, colour = "Estimation")) + 
ylab(TeX("$\\beta_j$ vs $\\hat{\\beta_j}$")) + 
theme(legend.title = element_blank())
```

Posterior probability of $\gamma_j=1$.

```{r}
gamma.sample <- res$sample.gamma[(burnin + 1):nMc, ]
post.gamma <- colMeans(gamma.sample)
qplot(x = 1:5, y = true.gamma, colour = "Truth", geom = "point") + 
  geom_point(aes(y = post.gamma, colour = "Estimation")) + 
  coord_cartesian(ylim = c(0, 5)) + 
  coord_cartesian(xlim = c(1, 5)) +
  ylab(TeX("$\\gamma_j$ vs $\\hat{\\gamma_j}$")) + 
  theme(legend.title = element_blank())
```

Posterior mean of $\sigma^2$.
 
```{r}
sigma2.sample <- res$sample.sigma2[(burnin + 1):nMc]
sig2.mean <- mean(sigma2.sample)  
qplot(sigma2.sample, geom = "histogram") + 
  geom_vline(xintercept = sig2.mean, color = "blue") +
  xlab(TeX("$\\hat{\\sigma^2}$ sample")) 
```

## 2. Example:  Bayesian variables selection with with spike-and-slab by Gibbs sampling algorithm.

We use the same settings from previous section. We need to re-estimate the parameters and reset the initial values.

```{r}
X0 <- covariate
beta.hat <- solve(t(X0) %*% X0) %*% t(X0) %*% Y
sig2.hat <- sum((Y - X0 %*% beta.hat) ** 2) / n

nMc <- 100
a <- (sig2.hat) ** 2 / 10 + 2  
b <- sig2.hat * (a - 1)
v0 <- rep(10e-3, p)
v1 <- rep(sqrt(max(beta.hat) * 100), p)
w <- 1 - sum(abs(beta.hat) < 4) / p
gamma0 <- rbinom(p, size = 1, prob = w)
sigma0 <- sig2.hat
```

Apply the algorithm and estimate compuation time.

```{r}
ptm <- proc.time()

res <- BvsGibbs(covariate, Y, nMc, v1, v0, w, gamma0, sigma0, a, b)

Time.used <- proc.time() - ptm
Time.used[3] / 60
```

The results would be close to the previous ones. So is the code.

```{r, echo=F, eval=F}
## Get posterior mean
burnin <- nMc / 2
beta.sample <- res$sample.beta[(burnin + 1):nMc, ]
beta.mean <- colMeans(beta.sample)
par(mfrow <- c(1, 1))

## Posterior inference of beta.
qplot(x = 1:5, y = true.beta, colour = "Truth", geom = "point") + 
  geom_point(aes(y = beta.mean, colour = "Estimation")) + 
  ylab(TeX("$\\beta_j$ vs $\\hat{\\beta_j}$")) + 
  theme(legend.title = element_blank())

## Posterior probability of gamma_j=1
gamma.sample <- res$sample.gamma[(burnin + 1):nMc, ]
post.gamma <- colMeans(gamma.sample)
qplot(x = 1:5, y = true.gamma, colour = "Truth", geom = "point") + 
  geom_point(aes(y = post.gamma, colour = "Estimation")) + 
  coord_cartesian(ylim = c(0, 5)) + 
  coord_cartesian(xlim = c(1, 5)) +
  ylab(TeX("$\\gamma_j$ vs $\\hat{\\gamma_j}$")) + 
  theme(legend.title = element_blank())

## Posterior mean of sigma2
sigma2.sample <- res$sample.sigma2[(burnin + 1):nMc]
sig2.mean <- mean(sigma2.sample)  
qplot(sigma2.sample, geom = "histogram") + 
  geom_vline(xintercept = sig2.mean, color = "blue") +
  xlab(TeX("$\\hat{\\sigma^2}$ sample"))  
```

## 3. Example: Bayesian variables selection with multiple linear regression by Metropolis–Hastings algorithm.

We use the same settings from previous section. We need to re-estimate the parameters and reset the initial values.

```{r}
X0 <- cbind(1, covariate)
beta.hat <- solve(t(X0) %*% X0) %*% t(X0) %*% Y
sig2.hat <- sum((Y - X0 %*% beta.hat) ** 2) / n

nMc <- 100
a <- (sig2.hat) ** 2 / 10 + 2 
b <- sig2.hat * (a - 1)
v0 <- rep(10e-6, p)
v1 <- rep(sqrt(max(beta.hat) * 100), p) / sqrt(sig2.hat)
w <- 1 - sum(abs(beta.hat) < 4) / p
gamma0 <- rep(0, p)
```

Apply the algorithm and estimate compuation time.

```{r}
ptm <- proc.time()

res <- BvsMH(covariate, Y, nMc, v1, v0, w, gamma0, a, b, phi)

Time.used <- proc.time() - ptm
Time.used[3] / 60
```

Do Posterior inference of $\boldsymbol\gamma$.

```{r}
burnin <- 30
accep.rate <- res$accept.count / nMc
gamma.sample <- res$sample.gamma[(burnin + 1):nMc, ]
post.gamma <- colMeans(gamma.sample)

qplot(x = 1:5, y = true.gamma, colour = "Truth", geom = "point") + 
  geom_point(aes(y = post.gamma, colour = "Estimation")) + 
  coord_cartesian(ylim = c(0, 5)) + 
  coord_cartesian(xlim = c(1, 5)) + 
  xlab("x") + 
  ylab("posterior gamma") + 
  ggtitle(TeX("marginal posterior prob of $\\gamma_j=1$"))
```

## 4. Example: Solving Lasso by ADMM.

Read data.

```{r}
n <- 500                 
p <- 1000         
maxiter <- 1000
```

The number of inportant features:

```{r}
s <- 100       
```

Coefficient:

```{r}
beta <- c(runif(s, -2, 2), rep(0, p - s))
A <- matrix(rnorm(p*p), p, p)
A <- scale(A, TRUE, TRUE)
b <-  A %*% beta + 0.1*rnorm(n)
```

Tunning paramter:

```{r}
lambda <- 0.1*max(abs(t(A) %*% b))
```

Term will be used:

```{r}
rho <- 0.5
solATA <- solve(t(A) %*% A + rho*diag(p))
ATb <- t(A) %*% b
```

Run the solver.

```{r}
rrs <- ADMM(rho, solATA, ATb, p, maxiter)
rrs.fast <- ADMM.fast(rho, solATA, ATb, maxiter)
```

Combine data.

```{r}
data <- data.frame(iter = seq(maxiter), rrs = rrs, rrs.fast = rrs.fast)
```

Gather data (gather in tidyr) or melt data (melt in reshape2).

```{r}
test.data <- melt(data, id = "iter")
ggplot(test.data, aes(x = iter, y = value, colour = variable)) + geom_line()
```

## 5. Example: Lasso path by using 'prostateStnd' data

Load data from `pmtk3` .

```{r}
data <- readMat('data/prostateStnd.mat')
X <- data$Xtrain
Y <- data$ytrain
```

Fit lasso:

```{r}
model <- glmnet(X, Y)
```

Extract lambda and estimated coefficient:

```{r}
lambda <- model$lambda
beta.hat <- as.matrix(model$beta)
data <- data.frame(lambda, t(beta.hat))
names(data) <- c("lambda", "V1", "V2", "V3", "V4", "V5", "V6", "V7", "V8")
data <- melt(data, c("lambda"))
```

Regularization path:

```{r}
ggplot(data, aes(x = lambda,y = value)) +
  geom_point(aes(colour = variable), size = .5) +
  geom_line(aes(colour = variable)) + 
  scale_x_reverse() + 
  coord_cartesian(xlim = c(0, max(lambda) + 2 * sd(lambda))) +
  coord_cartesian(ylim = c(min(beta.hat) - 2 * sd(beta.hat), 
                           max(beta.hat) + 2 * sd(beta.hat))) +
  xlab(TeX("$\\lambda$")) + 
  ylab(TeX("$\\hat{w}_j(\\lambda)$"))
```

## 6. Example: Ridge path by using 'prostateStnd' data

Fit ridge regression with same data from previous session:

```{r}
lambda <- seq(0, 100, 2)
beta.hat <- matrix(NA, nrow = length(lambda), ncol = dim(X)[2])
for (i in 1:length(lambda)){
  beta.hat[i, ] <- solve(t(X) %*% X + lambda[i] * diag(dim(X)[2])) %*% t(X) %*% Y
}
data <- data.frame(lambda, beta.hat)
names(data) <- c("lambda", "V1", "V2", "V3", "V4", "V5", "V6", "V7", "V8")
data <- melt(data, c("lambda"))
```

Regularization path:

```{r}
ggplot(data, aes(x = lambda, y=  value)) +
  geom_point(aes(colour = variable), size = .5) +
  geom_line(aes(colour = variable)) + 
  scale_x_reverse() + 
  coord_cartesian(xlim = c(0, max(lambda) + 2 * sd(lambda))) +
  coord_cartesian(ylim = c(min(beta.hat) - 2 * sd(beta.hat), 
                           max(beta.hat) + 2 * sd(beta.hat))) +
  xlab(TeX("$\\lambda$")) + 
  ylab(TeX("$\\hat{w}_j(\\lambda)$"))
```