---

title: 'Chapter 21: Variational inference'
author: "Jiali Lin"
date: "5/3/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(SML)
library(ggplot2)
library(mvtnorm)
if(!require("mixtools")) { install.packages("mixtools");  require("mixtools") }
```

## 1. Example: Illustrating forwards vs reverse KL on a bimodal distribution

Generate Data.

```{r}
mu <- matrix(c(-1, -1, 1, 1), nrow = 2,byrow = T);

Sigma <- array(0,c(2, 2, 2));
Sigma[, , 1] <- matrix(c(1/2, 1/4, 1/4, 1), nrow = 2);
Sigma[, , 2] <- matrix(c(1/2, -1/4, -1/4, 1), nrow = 2);
SigmaKL <- matrix(c(3, 2, 2, 3), nrow = 2);

x1 <- seq(-4, 4, 0.1);
x2 <- x1;

n1 <- length(x1);
n2 <- length(x2);

f1 <- matrix(0, nrow = n1, ncol = n2); 
f2 <- matrix(0, nrow = n1, ncol = n2); 
klf <- matrix(0, nrow = n1, ncol = n2); 
kll <- klf;
klr <- klf;

for (i in 1:n1){
  f1[i,] <- dmvnorm(cbind(rep(x1[i], n2), x2), mu[1, ], Sigma[, , 1]);
  f2[i,] <- dmvnorm(cbind(rep(x1[i], n2), x2), mu[2, ], Sigma[, , 2]);
  klf[i,] <- dmvnorm(cbind(rep(x1[i], n2), x2), matrix(c(0, 0), ncol = 1), SigmaKL);
  kll[i,] <- dmvnorm(cbind(rep(x1[i], n2), x2), mu[1, ], Sigma[, , 1] * 0.6);
  klr[i,] <- dmvnorm(cbind(rep(x1[i], n2), x2), mu[2, ], Sigma[, , 2]*  0.6);
}
```

Average two densities:

```{r}
f <- f1 + f2;
```

Make figures:

```{r}
contour(x1, x2, f, col = "blue")
contour(x1, x2, klf, add = TRUE, col = "red") 

contour(x1, x2, f, col = "blue")
contour(x1, x2, kll, add = TRUE, col = "red") 

contour(x1, x2, f, col = "blue")
contour(x1, x2, klr, add = TRUE, col = "red") 
```

## 2. Example: Illustrating forwards vs reverse KL on a symmetric Gaussian

Generate Data.

```{r}
mu <- c(0, 0)

Sigma <- matrix(c(1, 0.97, 0.97, 1), ncol = 2);
SigmaKLa <- diag(2)/25;
SigmaKLb <- diag(2);


x1 <- seq(-1, 1, 0.01);
x2 <- x1;

n1 <- length(x1);
n2 <- length(x2);

f <- matrix(NA, nrow = n1, ncol = n2); 
klqp <- f;
klpq <- f;  

for (i in 1:n1){
  f[i, ] <- dmvnorm(cbind(rep(x1[i], n2), x2), mu, Sigma);
  klqp[i, ] <- dmvnorm(cbind(rep(x1[i], n2), x2), mu, SigmaKLa);
  klpq[i, ] <- dmvnorm(cbind(rep(x1[i], n2), x2), mu, SigmaKLb);
}
```

Make figures.

```{r}
contour(x1, x2, f, col = "blue")
contour(x1, x2, klqp, add = TRUE, col = "red") 

contour(x1, x2, f, col = "blue")
contour(x1, x2, klpq, add = TRUE, col = "red") 
```

## 3. Example: Variational bayes for univariate Gaussian

Simulate data:

```{r}
set.seed(12) 
N <- 10  
D <- 1 
data <- runif(N) 
data <- (data - mean(data)) / sd(data) 
data <- matrix(data) 
```

Pick hyper parameters:

```{r}
prior <- list()
prior$a0 <- 0
prior$b0 <- 0
prior$mu0 <- 0
prior$kappa0 <- 0
```

Initialize VB to fit univariate gaussian:

```{r}
post <- list()
post$aN <- 2.5  
post$bN <- 1 
post$muN <- 0.5  
post$kappaN <- 5 
```

Set the true posterior model using normal gamma prior
```{r}
truePost <- list()
truePost$mu <- mean(data) 
truePost$kappa <- N 
truePost$alpha <- N/2 
truePost$beta <- 1/2*sum((data - mean(data) ) ** 2) 
```

Obtain target distribution:

```{r}
mu <- seq(-1.0 , 1.0 , by = 0.1)
lambda <- seq(0, 2, by = 0.1)
dens <- outer(mu, lambda)

## Plot truth
par(mfrow = c(2, 2))
contour(mu, lambda, dens, 
        xlab = 'mu', ylab = 'lambda', col = 'green', drawlabels = F)
```

Set options:

```{r}
Options <- list()
Options$maxIter <- 100
Options$iter <- 1
Options$Lq <- 0
Options$converged <- FALSE
Options$tol <- 1e-5  
Options$Lbound <- rep(NA, Options$maxIter)  
Options$Lbound[1] <- Options$Lq
```

Define true joint distribution:

```{r}
TrueNormalGammaPdf <- function(mu, lambda){
  muprior <- truePost$mu  
  kappa <- truePost$kappa   
  alpha <- truePost$alpha  
  beta <- truePost$beta 
  C <- (beta ** alpha * sqrt(kappa)) / (gamma(alpha) * sqrt(2 * pi))
  p <- C * (lambda ** (alpha-1/2)) * (exp(-beta * lambda)) *  
    (exp(-kappa / 2 * (lambda * (mu - muprior) ** 2)))
  return(p)
}
```

Define approximate of joint normal-gmma density. 

```{r}
vbPost <- function(mu, lambda){
  C <- (bN ** aN * sqrt(kappaN)) / (gamma(aN) * sqrt(2 * pi))
  p <- C * (lambda ** (aN-1/2)) * (exp(-bN * lambda)) *  
    (exp(-kappaN/2* (lambda * (mu - muN) ** 2)))
  return(p)
}
```


Run the Variational Bayes.

```{r}
res = UniGaussVb(data, prior, post, Options, TrueNormalGammaPdf, vbPost)
```

## 4. Example: multivariate Variational EM algorithm

Read faithful (2-d) data set and make plot.

```{r}
X <- faithful
plot(X)
```

Standardize the data.

```{r}
X <- t(X)
X <- X - kronecker(matrix(1, 1, ncol(X)),apply(X, 1, mean))	
X <- X / kronecker(matrix(1, 1, ncol(X)),apply(X, 1, var))	
dim <- dim(X)[1]
N <- dim(X)[2]
```

Initialize with EM algorithm.

```{r}
ncentres <- 15;
mix <- list();
mix$ncentres <- ncentres;
mix$priors <-  matrix(1, 1, mix$ncentres) / mix$ncentres;
mix$centres <- matrix(rnorm(mix$ncentres * dim, mean = 0, sd = 1),  
                      mix$ncentres, dim);
mix$covars <- kronecker(array(1, c(1, 1, mix$ncentres)), diag(dim));	
```

Intialize the priors.

```{r}
PriorPar <- list();
PriorPar$alpha <- .001;
PriorPar$mu <- matrix(0, dim, 1);
PriorPar$beta <- 1;
PriorPar$W <- 200 * diag(dim);
PriorPar$v <- 20;
```

Set the options for VBEM.

```{r}
options <- list();
options$maxIter <- 100;
options$threshold <- 1e-5;
options$displayFig <- TRUE;
options$displayIter <- TRUE;
```

Call the function:
  
```{r}
res <- GmmVbem(X, mix, PriorPar, options)
```


Plot lower bound.

```{r}
plot(res$L, lty = 2, type ="l", 
     ylab = "lower bound on log marginal likelihood",
     main = "variational Bayes objective for GMM")
```